---
title: "Class05: Data Vis with ggplot"
author: "Alyssa (PID: A16217030)"
format: pdf
---

# Graphics Systems in R 

There are many graphics systems in R for making plats and figures 

We have already played a little with **"base R"** graphics and the `plot()` function

Today we will start learning about a popular graphics package called `ggplot2()`

This is an add on package - i.e. we need to install it. I install (like I install any package) with the `install.packages()` function.

```{r}
plot(cars)
```

Before I can use the function from a package I have to load up the package from my "library". We use the `library` command to load it up 

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is made up of least 3 things: 
  - data (the nummbers etc. that will go into your plot)
  - aes (how the columns of data map to the plot aesthetics)
  - geoms (how the actually loooks, points, bars, lines, etc.)
  
```{r}
p <- ggplot(cars) + aes(x=speed, y=dist) + geom_point()
p
```
```{r}
library(patchwork)

(p | p / p)
```
  
  
  For simple plots ggplot is more verbose - it takes more code _ than based R plots.
  
  Add some more layers  to our ggplot
```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth(method= "lm", se=FALSE) + labs(title= "stopping distance of old cars ", subtitle = "A silly example plot") + theme_bw()
```
  
  
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
nrow
colnames(genes)
ncol(genes)
table(genes$State)
round(table(genes$State)/nrow(genes) * 100, 2)
ggplot(genes) + aes(x= Condition1, y= Condition2, col=State) + geom_point() + scale_colour_manual( values=c("blue", "grey", "red")) + labs( title= "Gene Expresion Changes Upon Drug Treatment", x= "Control (no drug)", y= "Drug Treatment" )


```
```{r}

```

  
```{r}
# install.packages("dplyr")  ## un-comment to install if needed
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)

ggplot(gapminder_2007) + aes( x=gdpPercap, y=lifeExp, color= pop) + geom_point(alpha= 0.5)
```

```{r}
ggplot(gapminder_2007) + aes( x=gdpPercap, y=lifeExp, size= pop) + geom_point(alpha= 0.8)
```
  
```{r}
ggplot(gapminder_2007) + geom_point(aes( x=gdpPercap, y=lifeExp, size= pop), alpha= .5) + scale_size_area(max_size=10)
```
  
```{r}
gapminder <- read.delim(url)
library(dplyr)

gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + geom_point(aes( x=gdpPercap, y=lifeExp, size= pop, color= continent), alpha= .7) + scale_size_area(max_size=15)
```
```{r}
ggplot(gapminder_2007) + geom_point(aes( x=gdpPercap, y=lifeExp, size= pop, color= continent), alpha= .7) + scale_size_area(max_size=15)
```
```{r}
ggplot(gapminder_2007) + geom_point(aes( x=gdpPercap, y=lifeExp, size= pop, color= continent), alpha= .7) + scale_size_area(max_size=15)
```
  
  